A Type System for Structured Aliasing in Parallel Programs

### 2012-10-30 doc/types-pldi2013/intro.tex
minor edits

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system enables a scalable, hierarchical 
distributed scheduling algorithm that enables Legion programs to achieve
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that enable programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

>>>>

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system permits a scalable, hierarchical 
distributed scheduling algorithm that enables Legion programs to achieve
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that allows programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;
